---
title: "Capstone"
output: html_document
date: "2024-10-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load in the Dataset
```{r pressure, echo=FALSE}
tuition_data<-read.csv("/Users/jiawendavis/Downloads/Top Universities - Sheet1 (1).csv")
```

```{r}
# Load necessary library
library(tidyverse)

# Convert all tuition-related columns to numeric before pivoting
tuition_data_long <- tuition_data %>%
  # Convert all columns that start with "X20" to numeric to ensure consistent data type
  mutate(across(starts_with("X20"), as.numeric)) %>%
  # Select all columns up to "State" (position 6) and the tuition columns
  select(1:6, starts_with("X20")) %>%
  # Pivot tuition columns to long format
  pivot_longer(
    cols = starts_with("X20"),         # Select yearly tuition columns
    names_to = c("Year", "Tuition_Type"), # New columns for year and tuition type
    names_pattern = "X(\\d{4}\\.\\d{4})\\..*(In.State|OOS)", # Regex to split year and type
    values_to = "Tuition"             # Column for tuition values
  ) %>%
  # Clean up the Tuition_Type column and reformat Year
  mutate(
    Tuition_Type = recode(Tuition_Type, "In.State" = "In-State", "OOS" = "Out-of-State"),
    Year = gsub("\\.", "-", Year) # Replace '.' with '-' in Year for consistency
  )

tuition_data_long <- na.omit(tuition_data_long) #omits columns related to enrollment and net price

# View transformed data
head(tuition_data_long)

```

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)

# Assuming tuition_data_long is your transformed dataset from above code
# Group data by Public or Private, Year, and Tuition_Type and calculate mean tuition
means <- tuition_data_long %>%
  group_by(Public.or.Private, Year, Tuition_Type) %>%
  summarise(Mean_Tuition = mean(Tuition, na.rm = TRUE))

# Create Profile Plot
plot <- ggplot(means, aes(x = Year, y = Mean_Tuition, color = Public.or.Private, group = Public.or.Private)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  facet_wrap(~ Tuition_Type) +
  theme_minimal() +
  labs(title = "Profile Plot of Public vs. Private College Tuition",
       x = "Academic Year", y = "Average Tuition (USD)",
       color = "Institution Type") +
  scale_color_manual(values = c("Public" = "blue", "Private" = "red")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability

print(plot)
```

```{r}
# Load necessary libraries
library(tidyverse)

# Define Ivy League colleges
ivy_league <- c("Harvard", "Yale", "Princeton", "Columbia", 
                "Brown", "Dartmouth", "University of Pennsylvania", "Cornell")

# Filter data for Ivy League and Other Private schools
ivy_data <- tuition_data_long %>%
  filter(City %in% ivy_league) 

other_private_data <- tuition_data_long %>%
  filter(!(City %in% ivy_league) & Public.or.Private == "Private") 

# Create a boxplot to compare tuition
ggplot() +
  geom_boxplot(data = ivy_data, aes(x = "Ivy League", y = Tuition), fill = "lightblue", alpha = 0.6) +
  geom_boxplot(data = other_private_data, aes(x = "Other Private", y = Tuition), fill = "salmon", alpha = 0.6) +
  labs(title = "Tuition Comparison: Ivy League vs. Other Private Schools",
       x = "School Type", y = "Tuition") +
  theme_minimal()

# t-test to compare the means of the two groups
t_test_result <- t.test(ivy_data$Tuition, other_private_data$Tuition)

# Print the results of the t-test
print(t_test_result)

```
The p-value is less than the common alpha level of 0.05. This means we can reject the null hypothesis, which posits that there is no difference in mean tuition between the two groups. The data suggests there is a statistically significant difference.

The 95% confidence interval for the difference in means is between -7385.991 and -1634.531. This interval does not include zero, further supporting the conclusion that there is a significant difference in tuition. Specifically, we can say that, on average, Ivy League schools charge between approximately $1,634 and $7,386 less than other private institutions.

Mean of Ivy League (x): 49,332.00
Mean of Other Private (y): 53,842.26
This shows that the average tuition for Ivy League schools is approximately $4,510 less than for other private schools.

